---
title: "MultispeciesMSE"
author: "Sarah Gaichas"
date: "1/11/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(mskeyrun)
```

Steps for multispecies MSE:

#  Condition operating model on data

## Data

Simulated output of ecosystem model. Options include Norwegian-Barents Atlantis generated dataset in [`mskeyrun`](https://noaa-edab.github.io/ms-keyrun/) R package and generating data from the [Georges Bank Rpath model](https://github.com/NOAA-EDAB/GBRpath). 

### mskeyrun (NOBA Atlantis) data

Which species to use in the model? Probably the ones that interact the most. Here we use simulated survey diet composition to find the highest average diet proportions between the 11 species in the simulated dataset. We are assuming we only want to focus on managed fish species, rather than interactions with lower trophic levels or protected species. 

```{r}
# install data package if needed
#remotes::install_github("NOAA-EDAB/ms-keyrun")
# the libraries are all in the first code chunk above now
#library(mskeyrun)

# which species pairs interact the most? 
# a too-quick analysis finding the mean proportion in diet from the diet dataset:

predprey <- mskeyrun::simSurveyDietcomp %>%
  filter(prey %in% unique(Name)) %>%
  select(Name, agecl, year, prey, value) %>%
  group_by(Name, prey) %>%
  summarise(avgpreyprop = mean(value)) %>%
  unite('Pred-Prey', Name:prey, sep = "-")

predprey %>% arrange(desc(avgpreyprop))

```

Select species from the mskeyrun dataset and get survey biomass and catch data for fitting. For now, pull Redfish and Blue_whiting. Later could add North_atl_cod, Capelin, Haddock as possibilities. 

```{r}

index.df <- mskeyrun::simSurveyIndex %>% filter(Name %in% c("Redfish", "Blue_whiting"))
harvest.df <- mskeyrun::simCatchIndex %>% filter(Name %in% c("Redfish", "Blue_whiting"))

data.years <- unique(mskeyrun::simSurveyIndex$year)

```

Plot them

```{r}
# plot(data.years,index, pch=19,xlab="Year",ylab="Million tonnes (B or C)",
#      ylim=c(0,1200))
# lines(data.years,harvest,lty=2,lwd=2)

index.df %>%
  filter(variable=="biomass") %>%
  ggplot() +
  geom_point(aes(x=year, y=value, colour=survey)) +
  geom_line(data=(harvest.df %>% filter(variable=="catch")), 
            aes(x=year, y=value)) +
  theme_bw() +
  facet_wrap(~Name)
  
```

Do we want to use just one of the surveys if we use this dataset? Maybe fall because Blue whiting migrate from the system in the spring (though maybe that is when Redfish eat them all? might be worth a look)

### Rpath data

Full dataset now in the `data` folder in this repo.
```{r}

#load Rpath Georges Bank model outputs, object is called GB.data
load(here("multispecies/data/GBdata.rda"))

#units are tons/square kilometer for biomass and catch
str(GB.data)

#there are three spexxies
unique(GB.data$Species)

data.years <- unique(GB.data$Year)

GB.data %>%
  ggplot()+
  geom_point(aes(x=Year, y=ObsBio)) +
  geom_line(aes(x=Year, y=TotCatch)) +
  theme_bw() +
  facet_wrap(~Species, scales = "free")

```


## Multispecies operating model

Production model or biomass-dynamics model form with species interaction terms. Use Gavin's `schaefer` function as modified below and reuse as many other functions as possible.

What interactions to model?

*  Predation mortality 
*  Prey impacts on growth?

### Schaefer model (single species, directly from [my-first-mse](https://github.com/gavinfay/cinar-mse/blob/main/materials/exercises/day-01/my-first-mse.Rmd))

First, the logistic production function:  
```{r}
schaefer <- function(B,C,K,r) {
  #function schaefer takes the current biomass, a catch, 
  #and the model parameters to compute next year's biomass
  res <- B + B * r * (1 - B/K) - C
  return(max(0.001,res))  # we add a constraint to prevent negative biomass
}
```

Now a function to do the biomass projection:  
```{r}
dynamics <- function(pars,C,yrs) {
  # dynamics takes the model parameters, the time series of catch, 
  # & the yrs to do the projection over
  
  # first extract the parameters from the pars vector (we estimate K in log-space)
  K <- exp(pars[1])
  r <- exp(pars[2])
  
  # find the total number of years
  nyr <- length(C) + 1
  
  # if the vector of years was not supplied we create 
  # a default to stop the program crashing
  if (missing(yrs)) yrs <- 1:nyr
  
  #set up the biomass vector
  B <- numeric(nyr)
  
  #intialize biomass at carrying capacity
  B[1] <- K
  # project the model forward using the schaefer model
  for (y in 2:nyr) {
    B[y] <- schaefer(B[y-1],C[y-1],K,r)
  }
  
  #return the time series of biomass
  return(B[yrs])
  
  #end function dynamics
}  
```


We are going to condition the operating model by estimating the parameters based on the historical biomass index data.  

To do this we make a function that shows how well the current parameters fit the data, we assume that the observation errors around the true biomass are log-normally distributed.  

```{r}
# function to calculate the negative log-likelihood
nll <- function(pars,C,U) {  #this function takes the parameters, the catches, and the index data
  sigma <- exp(pars[3])  # additional parameter, the standard deviation of the observation error
  B <- dynamics(pars,C)  #run the biomass dynamics for this set of parameters
  Uhat <- B   #calculate the predicted biomass index - here we assume an unbiased absolute biomass estimate
  output <- -sum(dnorm(log(U),log(Uhat),sigma,log=TRUE),na.rm=TRUE)   #calculate the negative log-likelihood
  return(output)
  #end function nll
}
```


Function to perform the assessment and estimate the operating model parameters  
(i.e. to fit the logistic model to abundance data)
```{r}
assess <- function(catch,index,calc.vcov=FALSE,pars.init) {
  # assess takes catch and index data, initial values for the parameters,
  # and a flag saying whether to compute uncertainty estimates for the model parameters
  
  #fit model
  # optim runs the function nll() repeatedly with differnt values for the parameters,
  # to find the values that give the best fit to the index data
  res <- optim(pars.init,nll,C=catch,U=index,hessian=TRUE)
  
  # store the output from the model fit
  output <- list()
  output$pars <- res$par
  output$biomass <- dynamics(res$par,catch)
  output$convergence <- res$convergence
  output$nll <- res$value
  if (calc.vcov)
    output$vcov <- solve(res$hessian)
  
  return(output)
  #end function assess
}
```


## Conditioning

### Single species models?

Try fitting to each species separately first, adjusting Gavin's code (test with atlantis output till we get rpath):

Make separate data objects for each species
```{r}
harvest.redfish <- harvest.df %>%
  filter(Name=="Redfish",
         variable=="catch") %>%
  select(value) %>%
  unlist() %>%
  unname()

index.redfish <- index.df %>%
  filter(Name=="Redfish",
         variable=="biomass", 
         survey=="BTS_fall_allbox_effic1") %>%
  select(value) %>%
  unlist() %>%
  unname()

harvest.bluewhiting <- harvest.df %>%
  filter(Name=="Blue_whiting",
         variable=="catch") %>%
  select(value) %>%
  unlist() %>%
  unname()

index.bluewhiting <- index.df %>%
  filter(Name=="Blue_whiting",
         variable=="biomass", 
         survey=="BTS_fall_allbox_effic1") %>%
  select(value) %>%
  unlist() %>%
  unname()

harvest.GBcod <- GB.data %>%
  filter(Species=="Cod") %>%
  select(TotCatch) %>%
  unlist() %>%
  unname()

index.GBcod <- GB.data %>%
  filter(Species=="Cod") %>%
  select(ObsBio) %>%
  unlist() %>%
  unname()

harvest.GBherring <- GB.data %>%
  filter(Species=="AtlHerring") %>%
  select(TotCatch) %>%
  unlist() %>%
  unname()

index.GBherring <- GB.data %>%
  filter(Species=="AtlHerring") %>%
  select(ObsBio) %>%
  unlist() %>%
  unname()

harvest.GBhaddock <- GB.data %>%
  filter(Species=="Haddock") %>%
  select(TotCatch) %>%
  unlist() %>%
  unname()

index.GBhaddock <- GB.data %>%
  filter(Species=="Haddock") %>%
  select(ObsBio) %>%
  unlist() %>%
  unname()  
  
```


First define initial parameter vector for each species: log(K), log(r), log(sigma)  
```{r}
#ini.parms <- c(log(1200), log(0.1), log(0.3))

# use max survey B as the initial value for K
maxsvB.red <- max(index.redfish)
ini.parms.red <- c(log(maxsvB.red), log(0.1), log(0.3))

maxsvB.blue <- max(index.bluewhiting)
ini.parms.blue <- c(log(maxsvB.blue), log(0.5), log(0.3))

maxsvB.GBcod <- max(index.GBcod)
ini.parms.GBcod <- c(log(maxsvB.GBcod), log(0.2), log(0.3))

maxsvB.GBherring <- max(index.GBherring)
ini.parms.GBherring <- c(log(maxsvB.GBherring), log(0.3), log(0.3))

maxsvB.GBhaddock <- max(index.GBhaddock)
ini.parms.GBhaddock <- c(log(maxsvB.GBhaddock), log(0.3), log(0.3))


```


Fit the logistic model to data (redfish):  
```{r}
redfish <- assess(harvest.redfish,index.redfish,calc.vcov=TRUE,ini.parms.red)
redfish
```

Fit the logistic model to data (bluewhiting):  
```{r}
bluewhiting <- assess(harvest.bluewhiting,index.bluewhiting,calc.vcov=TRUE,ini.parms.blue)
bluewhiting
```


SS model fit plots: fitted line in gray

```{r}

fit.red <- data.frame(value= redfish$biomass,
                      year=c(40:120),
                      Name="Redfish")

fit.blue <- data.frame(value= bluewhiting$biomass,
                       year=c(40:120),
                       Name="Blue_whiting")

fits <- bind_rows(fit.red, fit.blue)


index.df %>%
  filter(variable=="biomass") %>%
  ggplot() +
  geom_point(aes(x=year, y=value, colour=survey)) +
  geom_line(data=(harvest.df %>% filter(variable=="catch")), 
            aes(x=year, y=value)) +
  geom_line(data=fits, aes(x=year, y=value), lwd=1.3, col=gray(0.5)) +
  theme_bw() +
  facet_wrap(~Name)
```

Fit the logistic model to data (GBcod):  
```{r}
GBcod <- assess(harvest.GBcod,index.GBcod,calc.vcov=TRUE,ini.parms.GBcod)
GBcod
```

Fit the logistic model to data (GBherring):  
```{r}
GBherring <- assess(harvest.GBherring,index.GBherring,calc.vcov=TRUE,ini.parms.GBherring)
GBherring
```

Fit the logistic model to data (GBhaddock):  
```{r}
GBhaddock <- assess(harvest.GBhaddock,index.GBhaddock,calc.vcov=TRUE,ini.parms.GBhaddock)
GBhaddock
```

GB SS model fit plots: fitted line in gray
```{r}

#model may not be using the last data year? misaligned. a hack here to fix

fit.cod <- data.frame(value= GBcod$biomass,
                      Year=c(1983:2023),
                      Species="Cod")

fit.herring <- data.frame(value= GBherring$biomass,
                       Year=c(1983:2023),
                       Species="AtlHerring")

fit.haddock <- data.frame(value= GBhaddock$biomass,
                       Year=c(1983:2023),
                       Species="Haddock")

fits <- bind_rows(fit.cod, fit.herring, fit.haddock)


GB.data %>%
  ggplot()+
  geom_point(aes(x=Year, y=ObsBio)) +
  geom_line(aes(x=Year, y=TotCatch)) +
  geom_line(data=fits, aes(x=Year, y=value), lwd=1.3, col=gray(0.5)) +
  theme_bw() +
  facet_wrap(~Species, scales = "free")

```

#  Design HCRs

Separate single species HCRs (start with Gavin's specified as default)



#  Project with HCRs

#  Compare results

#  Tradeoffs